Typestate Inference in an Imperative Calculus

### 2012-11-05 formalism/paper.tex
addition of new join-of-separate-types type constraint to formalism (still needs full discussion).

@@

\[
\begin{tabular}{cc}
\infer
{\mu[x \mapsto v] = \mu, x \mapsto v}
{\mu = \overline{x_i \mapsto v_i} & x \notin \overline{x_i}}
&
\infer
{\mu[x \mapsto v] = \mu, x \mapsto v}
{\mu = \overline{x_i \mapsto v_i}, x \mapsto v'}
\end{tabular}
\]

As there is no way for a reference to be duplicated, the mapping
from variable names to store locations is injective and therefore we have
no need for explicit ``location values'' as in the typical treatment of 
languages with references (such as in \cite{Pierce2002}). There is no need
for a separate store typing, as a well-typed store can be defined directly
in relation to a context as in \figref{fig:storetype}. A well-typed store
may mappings for variables which are not in the context as shown in
rule \textsc{st\_extra}.

@@

\[
\begin{tabular}{cc}
\infer[\textsc{st\_empty}]
{\emptyset \vdash \emptyset}
{}
&
\infer[\textsc{st\_vars}]
{\overline{x_i : T_i} \vdash \overline{x_i \mapsto v_i}}
{\typerule{\Gamma}{v_i}{T_i}{\Gamma'}}
\\\\
\multicolumn{2}{c}{
\infer[\textsc{st\_extra}]
{\Gamma \vdash \mu, \overline{y_j \mapsto v_j}}
{\Gamma \vdash \mu & \overline{y_j \notin dom(\mu)}}
}
\end{tabular}
\]

@@

\begin{itemize}
\item $S \in O$, a predicate asserting $S$ is a state of $O$, is equivalent to
$S \in \Sigma$.
\item $m : T \Rightarrow S' \in O@S$, the method existence predicate, 
states the method $m$ with return type $T$ exists in state $S$ and triggers a 
transition to state $S'$ when called. Formally, $(S,m,T,S') \in \Delta$.
\item $m : T \Rightarrow \overline{S'} \in O@\overline{S}$ is the extension
of the previous to state sets, meaning that for all $S \in \overline{S}$
there exists a state $S'$ and type $T'$ where 
$(S,m,T',S') \in \Delta$ and that
$T = \bigsqcup_{S \in \overline{S}} \{ T' | (S,m,T',S') \in \Delta \}$ exists
and that $\overline{S'} = \{ S' | (S,m,T',S') \in \Delta \}$.
\item $O@\overline{S} \sim O_2@\overline{S_2}$, the extension of the
simulation relation to support state sets, can be interpreted as
as $\bigsqcup_{S \in \overline{S}}O@S \sim \bigsqcup_{S_2 \in \overline{S_2}}O_2@S_2$.
\end{itemize}

@@

\begin{figure}
\ottgrammartabular{
\ottC\ottinterrule
\ottctxc\ottinterrule
\otttyc\ottinterrule
\otttinf\ottinterrule
}
\caption{\label{fig:congrammar} Constraint grammar}
\end{figure}

\begin{figure}
\[
\begin{array}{l}

@@

% object value
\inferlhs{[ \overline{S_i \{ \overline{ m_{ij} = (v_{ij}, S_{ij}) } \} }]@S}
\triangleq
\exists \overline{\alpha_{ij}} . \\
\alpha = \{ \overline{ S_i \{ \overline{ m_{ij} : \alpha_{ij} >> S_{ij} } \} }\}@S
\cand
\gamma = \gamma'
\cand \\
\overline{\inferrule{\gamma}{\triangleright v_{ij}}{\alpha_{ij}}{\gamma}}
\\\\

% function value
\inferlhs{\lambda ( \overline{x_i} ) . t} \triangleq
\exists \gamma_r, \gamma_{in}, \gamma_{out}, \overline{\alpha_{ai}, \alpha_{bi}}, \alpha_r . \\
\alpha = (\overline{\alpha_{ai} \gg \alpha_{bi}}) \rightarrow \alpha_r
\cand
\gamma = \gamma'
\cand
\gamma_r = \emptyset
\cand \\
\gamma_{in} = \gamma_r, \overline{x_i : \alpha_{ai}}
\cand
\overline{x_i : \alpha_{bi} \in \gamma_{out}}
\cand
\\
\inferrule{\gamma_{in}}{t}{\alpha_r}{\gamma_{out}}
\\\\

% let binding
\inferlhs{\lett{x}{t'}{t}} \triangleq \exists \gamma_v, \gamma_{b1}, \gamma_{b2}, \alpha' . \\
\gamma_{b1} = \gamma_v, x : \alpha'
\cand
\gamma' = \gamma_{b2} - \{ x \}
\cand \\
\inferrule{\gamma}{t'}{\alpha'}{\gamma_v}
\cand
\inferrule{\gamma_{b1}}{t}{\alpha}{\gamma_{b2}}
\\\\

% sequence
\inferlhs{t ; t'} \triangleq
\exists \gamma'', \alpha' . \\
\inferrule{\gamma}{t}{\alpha'}{\gamma''}
\cand
\inferrule{\gamma''}{t'}{\alpha}{\gamma'}
\\\\

% function call
\inferlhs{x ( \overline{x_i} )} \triangleq
\exists \overline{\alpha_i}, \overline{\alpha_i'}, \overline{\alpha_i''} . \\
\overline{x_i : \alpha_i \in \gamma}
\cand 
x : ( \overline{\alpha_i' \gg \alpha_i''} ) \rightarrow \alpha \in \gamma
\cand
\\
\gamma' = \gamma [ \overline{x_i \mapsto \remapapp{\alpha_i}{\alpha_i'}{\alpha_i''} } ]
\\\\

% method call
\inferlhs{x.m} \triangleq
\exists \alpha_x . \\
x : \alpha_x \in \gamma
\cand 
\cand  \\
\gamma' = \gamma[ x \mapsto \remapapp{\alpha_x}{O@S}{O@S'} ]
\\
\mathrm{where } O = \{ S \{ m : \alpha \Rightarrow S' \} S' \{ \} \}
\\\\

% if-then-else
\inferlhs{\ift{t_c}{t_t}{t_f}} \triangleq
\\
\exists \alpha_c, \alpha_t, \alpha_f, \gamma_c, \gamma_t, \gamma_f . \\
\inferrule{\gamma}{t_c}{\alpha_c}{\gamma_c}
\cand
\inferrule{\gamma_c}{t_t}{\alpha_t}{\gamma_t}
\cand 
\\
\inferrule{\gamma_c}{t_f}{\alpha_f}{\gamma_f}
\cand
\alpha_c = \boolt
\cand
\\
\alpha = \alpha_t \sqcup \alpha_f
\cand
\gamma' = \gamma_t \sqcup \gamma_f
\\\\

>>>>

\DeclareSymbolFont{symbolsC}{U}{ntxsyc}{m}{n}
\SetSymbolFont{symbolsC}{bold}{U}{ntxsyc}{b}{n}
\DeclareMathSymbol{\sqcupplus}{\mathbin}{symbolsC}{124}

@@

\[
\begin{tabular}{cc}
\infer
{\mu[x \mapsto v] = \mu, x \mapsto v}
{\mu = \overline{x_i \mapsto v_i} & x \notin \overline{x_i}}
&
\infer
{\mu[x \mapsto v] = \overline{x_i \mapsto v_i} , x \mapsto v}
{\mu = \overline{x_i \mapsto v_i}, x \mapsto v'}
\end{tabular}
\]

As there is no way for a reference to be duplicated, the mapping
from variable names to store locations is injective and therefore we have
no need for explicit ``location values'' as in the typical treatment of 
languages with references (such as in \cite{Pierce2002}). There is no need
for a separate store typing, as a well-typed store can be defined directly
in relation to a context as in \figref{fig:storetype}. A well-typed store
may contain mappings for variables which are not in the context as shown in
rule \textsc{st\_extra}.

@@

\[
\begin{tabular}{cc}
\infer[\textsc{st\_empty}]
{\emptyset \vdash \emptyset}
{}
&
\infer[\textsc{st\_vars}]
{\overline{x_i : T_i} \vdash \overline{x_i \mapsto v_i}}
{\overline{\typerule{\Gamma}{v_i}{T_i}{\Gamma'}}}
\\\\
\multicolumn{2}{c}{
\infer[\textsc{st\_extra}]
{\Gamma \vdash \mu, \overline{x_i \mapsto v_i}}
{\Gamma \vdash \mu & \overline{x_i \notin dom(\mu)}}
}
\end{tabular}
\]

@@

\begin{itemize}
\item $S \in O$ is the {\it state existence predicate} asserting 
$S$ is a state of $O$. 
Formally, this is equivalent to $S \in \Sigma$.
\item $m : T \Rightarrow S' \in O@S$ is the {\it method existence predicate} 
asserting the existence
of method $m$ with return type $T$ in state $S$ which triggers a 
transition to state $S'$ when called (where both $S \in O$ and $S' \in O$). 
Formally, this is equivalent to $(S,m,T,S') \in \Delta$.
\item $m : T \Rightarrow \overline{S'} \in O@\overline{S}$ is the extension
of the method existence predicate to state sets, meaning that for all 
$S \in \overline{S}$ there exists a state and type 
$T'$ where $(S,m,T',S') \in \Delta$ and that
$T = \bigsqcup_{S \in \overline{S}} \{ T' | (S,m,T',S') \in \Delta \}$ exists
and that $\overline{S'} = \{ S' | (S,m,T',S') \in \Delta \}$.
\item $O@\overline{S} \sim O_2@\overline{S_2}$ is the extension of the
simulation relation to support state sets, and can be interpreted as
$\bigsqcup_{S \in \overline{S}}O@S \sim \bigsqcup_{S_2 \in \overline{S_2}}O_2@S_2$.
\end{itemize}

@@

\begin{figure}
\ottgrammartabular{
\ottC\ottinterrule
\ottctxc\ottinterrule
\otttyc\ottinterrule
\otttinf\ottinterrule
\ottoinf\ottinterrule
\ottsinf\ottinterrule
}
\caption{\label{fig:congrammar} Constraint grammar}
\end{figure}

\begin{figure}
\[
\begin{array}{ll}

@@

% object value
\inferlhs{[ \overline{S_i \{ \overline{ m_{ij} = (v_{ij}, S_{ij}) } \} }]@\overline{S}}
\triangleq \\
\exists \alpha_o, \alpha_s, \overline{\alpha_{ij}}, \gamma_{\emptyset} . \\
\gamma_{\emptyset} = \emptyset
\cand
\alpha = \alpha_o@\alpha_s
\cand
\overline{\inferrule{\gamma_{\emptyset}}{v_{ij}}{\alpha_{ij}}{\gamma_{\emptyset}}}
\cand
\\
\gamma = \gamma'
\cand
\alpha_o @ \alpha_s = \{ \overline{ S_i \{ \overline{ m_{ij} : \alpha_{ij} >> S_{ij} } \} }\} @ \overline{S}
\\\\

% function value
\inferlhs{\lambda ( \overline{x_i : T_i \gg T_i'} ) . t} \triangleq \\
\exists \gamma_r, \gamma_{in}, \gamma_{out}, \overline{\alpha_i, \alpha_i'}, \alpha_r . \\
\alpha = (\overline{\alpha_i \gg \alpha_i'}) \rightarrow \alpha_r
\cand
\gamma = \gamma'
\cand
\gamma_r = \emptyset
\cand \\
\gamma_{in} = \gamma_r, \overline{x_i : \alpha_i}
\cand
\overline{x_i : \alpha_i' \in \gamma_{out}}
\cand
\\
\inferrule{\gamma_{in}}{t}{\alpha_r}{\gamma_{out}}
\\
\cand
\overline{T_i <: \alpha_i}
\cand
\overline{\alpha_i' <: T_i'}
\\\\

% let binding
\inferlhs{\lett{x}{t'}{t}} \triangleq \\
\exists \gamma_v, \gamma_{b1}, \gamma_{b2}, \alpha' . \\
\gamma_{b1} = \gamma_v, x : \alpha'
\cand
\gamma' = \gamma_{b2} - \{ x \}
\cand \\
\inferrule{\gamma}{t'}{\alpha'}{\gamma_v}
\cand
\inferrule{\gamma_{b1}}{t}{\alpha}{\gamma_{b2}}
\\\\

% sequence
\inferlhs{t ; t'} \triangleq \\
\exists \gamma'', \alpha' . \\
\inferrule{\gamma}{t}{\alpha'}{\gamma''}
\cand
\inferrule{\gamma''}{t'}{\alpha}{\gamma'}
\\\\

% function call
\inferlhs{x ( \overline{x_i} )} \triangleq \\
\exists \overline{\alpha_i}, \overline{\alpha_i'}, \overline{\alpha_i''} . \\
\overline{x_i : \alpha_i \in \gamma}
\cand 
x : ( \overline{\alpha_i' \gg \alpha_i''} ) \rightarrow \alpha \in \gamma
\cand
\\
\gamma' = \gamma [ \overline{x_i \mapsto \remapapp{\alpha_i}{\alpha_i'}{\alpha_i''} } ]
\\\\

% method call
\inferlhs{x.m} \triangleq \\
\exists \alpha_o, \alpha_s . \\
x : \alpha_o@\alpha_s \in \gamma
\cand \\
\gamma' = \gamma[ x \mapsto \remapapp{\alpha_o@\alpha_s}{O@S}{O@S'} ]
\\
\mathbf{where}\; O = \{ S \{ m : \alpha \Rightarrow S' \} S' \{ \} \}
\\\\

% if-then-else
\inferlhs{\ift{t_c}{t_t}{t_f}} \triangleq \\
\exists \alpha_c, \alpha_t, \alpha_f, \gamma_c, \gamma_t, \gamma_f . \\
\inferrule{\gamma}{t_c}{\alpha_c}{\gamma_c}
\cand
\inferrule{\gamma_c}{t_t}{\alpha_t}{\gamma_t}
\cand 
\\
\inferrule{\gamma_c}{t_f}{\alpha_f}{\gamma_f}
\cand
\alpha_c = \boolt
\cand
\\
\alpha = \alpha_t \sqcupplus \alpha_f
\cand
\gamma' = \gamma_t \sqcup \gamma_f
\\\\
